---
title: "SARS-CoV-2 example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SARS-CoV-2 example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


In this vignette, we show an example of how to modify the arguments to simulate
SARS-CoV-2 S gene evolution. Note that this is purely an example and we are not experts in this area, 
so if you would like to simulate SARS-CoV-2 evolution for a research project, we would 
recommend looking into the specific parameter values yourself.

If you would like more detail about how to generate inputs or more details
about the `run_wavess()` function, please see `vignette("prepare_input_data")`
and `vignette("run_wavess")`, respectively. 


## Load libraries

First, we have to load the wavess library and a few additional libraries that
we'll use for this vignette:

```{r}
library(wavess)
library(ape)
library(dplyr)
library(tibble)
library(readr)
library(tidyr)
```

## Create Python virtual environment

Next, we must create a Python virtual environment. The virtual environment is
required because the underlying code of `run_wavess()` is written in Python. You
only have to create the virtual environment once on each machine.

```{r}
create_python_venv()
```

## Prepare input data

Next, we must generate the input data. If you're interested in learning more
about how to prepare the input data, please see the corresponding vignette (see
`vignette("prepare_input_data")`).

We will only simulate 60 generations with sampling every day so the
simulation doesn't take too long to run.

```{r setup}
pop <- define_growth_curve(n_gens = 60 / 0.125)
samp <- define_sampling_scheme(sampling_frequency_active = 1, sampling_frequency_latent = 0) |> filter(day <= 60)
```

Let's load in the founder sequence first:

```{r}
fa_file <- system.file("extdata", "EPI_ISL_402124-S.fasta", package = "wavess")
founder <- extract_seqs(read.dna(fa_file, format = "fasta"),
  founder = "EPI_ISL_402124-S"
)
```

We will need to change the generation time, mutation rate, and recombination rate.

Unlike HIV, SARS-CoV-2 has no latent reservoir, so we will 
turn this feature off in the model.

generation time - 3 hours --> 0.125 days  
https://www.pnas.org/doi/10.1073/pnas.2406303121

mutation rate - 1.5e-6 per site per generation  
https://www.nature.com/articles/s41467-025-61555-x

recombination rate - 2e-6 per site per year --> 2e-6/(365/0.125) = 6.8e-10 per site per generation  
https://www.nature.com/articles/s41467-022-31749-8

```{r}
gen_time <- 0.125
mut_rate <- 1.5e-6
rec_rate <- 6.8e-10
```


For simplicity, we will not model conserved sites or replicative fitness, and
we will use a Jukes Cantor model of nucleotide subtitution rather than a
SARS-CoV-2 specific Q matrix, although these would make the simulation output 
more realistic.

```{r}
q_mat <- matrix(
  c(
    -0.75, 0.25, 0.25, 0.25,
    0.25, -0.75, 0.25, 0.25,
    0.25, 0.25, -0.75, 0.25,
    0.25, 0.25, 0.25, -0.75
  ),
  ncol = 4
)
rownames(q_mat) <- c("A", "C", "G", "T")
colnames(q_mat) <- c("A", "C", "G", "T")
```


To get CTL epitopes for the founder amino acid sequence, we used the IEDB T-cell prediction 
tool for MHC class I molecules with HLA alleles A*02:01, A*01:01, B*07:02, B*08:01 
and pMHC immunogenicity prediction, both with the default settings. We filtered
them in the same way as we do for the HIV CTL epitopes.
Website: https://nextgen-tools.iedb.org/pipeline?tool=tc1

```{r}
# get the amino acid sequence for use with IEDB
founder_aa <- ape::trans(ape::as.DNAbin(strsplit(founder$founder, split = "")[[1]]))
```


```{r}
peptide_table_file <- system.file("extdata", "sars-cov-2-spike_peptide_table.tsv", package = "wavess")
ctl_epitopes <- read_tsv(peptide_table_file) %>%
  # arrange(`netmhcpan_el percentile`)
  filter(`netmhcpan_el percentile` < 0.1 & `immunogenicity score` > 1 / 90) %>%
  mutate(
    start = (start - 1) * 3,
    start = ifelse(`seq #` == 1, start, start + nchar(founder_ref_pol$founder)),
    days_to_full_potency = 1 / `immunogenicity score`
  ) %>%
  group_by(`seq #`, peptide, start) %>%
  summarize(days_to_full_potency = round(min(days_to_full_potency))) %>% # if multiple HLAs recognize the same epitope
  arrange(`seq #`, start) %>%
  rowwise() %>%
  mutate(
    anchor2 = strsplit(peptide, "")[[1]][2],
    anchor9 = strsplit(peptide, "")[[1]][9]
  ) %>%
  ungroup() %>%
  pivot_longer(starts_with("anchor"), names_to = "escape_position", values_to = "recognized_aa") %>%
  mutate(escape_position = as.numeric(gsub("anchor", "", escape_position))) %>%
  select(-c(peptide, `seq #`))
```

We downloaded SARS-CoV-2 spike protein antibody epitopes from the LANL database and 
used them in the same way as we use the HIV antibody epitopes.  
Website (Spike feature download): https://cov.lanl.gov/components/sequence/COV/neutralization/download_db.comp

```{r}
antibody_file <- system.file("extdata", "sars-cov-2-spike-antibody-features.txt", package = "wavess")
antibody_epitopes <- read_tsv(antibody_file) %>%
  pull(Position) %>%
  get_epitope_frequencies() %>%
  sample_epitopes()
```


## SARS-CoV-2 simulation with `run_wavess()`


```{r}
spike_sim <- run_wavess(
  inf_pop_size = pop,
  samp_scheme = samp,
  generation_time = gen_time,
  mut_rate = mut_rate,
  q = q_mat,
  recomb_rate = rec_rate,
  founder_seqs = rep(founder$founder, 10),
  b_epitope_locations = antibody_epitopes,
  t_epitope_locations = ctl_epitopes,
  act_to_lat = 0 # no latency
)
```

Summary:

```{r}
spike_sim$counts %>%
  select(-c(contains("latent"), contains("replicative"), contains("conserved"))) # not modeling these so they are uninteresting
```

Fitness:

```{r}
spike_sim$fitness
```

Sequences from active cells:

```{r}
spike_sim$seqs_active
```

If you'd like to learn more about how to analyze the output, please check out 
the post-processing vignette `vignette("analyze_output")`.

